PHQ-4-R - Data Cleaning (Study 2)

### 2024-12-07 study2/analysis/1_cleaning.qmd
paper

---
title: "PHQ-4-R - Data Cleaning (Study 2)"
editor: source
editor_options: 
  chunk_output_type: console
format:
  html:
    code-fold: true
    self-contained: true
---

@@

We removed `r length(outliers$attentionchecks)` participants for failing the attention checks.

@@

### Response Coherence

@@

df$BDI2_SD <- select(df, matches("BDI2_"), -BDI2_Duration, -BDI2_Order, -BDI2_Total) |> 
  SD_per_dim(dims=c("BDI2_")) |> 
  rowMeans()
df$STAI5_SD <- select(df, matches("STAI5_"), -STAI5_Duration, -STAI5_Order, -STAI5_General) |> 
  SD_per_dim(dims=c("STAI5_")) |> 
  rowMeans()
```

mgcv::gam(Failed ~ s(STAI5_SD), 
          data=mutate(df, Failed = ifelse(AttentionCheck == "Failed", 1, 0)), 
          family="binomial") |>
  parameters::parameters()

df |> 
  select(Participant, STAI5_SD, AttentionCheck) |> 
  mutate(Participant = fct_reorder(Participant, STAI5_SD)) |>
  ggplot(aes(x=Participant, y=STAI5_SD, fill=AttentionCheck)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  theme(axis.text.x = element_blank())
```

@@

```{r}
dfoutlier <- df |> 
  select(contains("IAS_"), -IAS_Duration, -IAS_Order, -IAS_Total) |>
  # select(contains("BDI2_"), -BDI2_Duration, -BDI2_Order, -BDI2_Total, -BDI2_SD) |> 
  # datawizard::remove_empty_rows() |> 
  performance::check_outliers(
    method=c("optics", "mahalanobis_robust"),
    threshold=list(optics=2.5, optics_xi=0.03)) |> 
  as.data.frame() |> 
  mutate(Participant = fct_reorder(df$Participant, Distance_OPTICS),
         Outlier_AttentionCheck = ifelse(Participant %in% outliers$attentionchecks, 1, 0),
         Outlier_Distance = ifelse(abs(standardize(Distance_OPTICS)) > qnorm(0.995), 1, 0),
         Experiment_Duration = df$Experiment_Duration,
         Outlier = ifelse(Outlier_AttentionCheck == 1, "Failed Attention Checks", "Passed"))

outliers$distance <- as.character(dfoutlier[dfoutlier$Outlier_Distance == 1, "Participant"])
outliers$distance <- outliers$distance[!outliers$distance %in% c(outliers$attentionchecks, outliers$duration)]

@@

```{r}
df <- filter(df, !Participant %in% c(outliers$attentionchecks, outliers$distance))
df <- select(df, -starts_with("AttentionCheck"))
```

@@

```{r}
p1 <- select(df, Participant, Gender, starts_with("Disorder_")) |> 
  pivot_longer(cols = starts_with("Disorder_"), names_to = "Disorder", values_to = "Value") |> 
  mutate(Disorder = str_remove_all(Disorder, fixed("Disorder_"))) |> 
  summarize(N = sum(Value) / nrow(df), .by=c("Gender", "Disorder")) |> 
  mutate(N_tot = sum(N), .by="Disorder") |>
  mutate(Disorder = fct_reorder(Disorder, desc(N_tot))) |>
  ggplot(aes(x = Disorder, y = N, fill=Gender)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Male"= "#64B5F6", "Female"= "#F06292", "Other"="orange", "Missing"="brown")) +
  scale_y_continuous(expand = c(0, 0), labels=scales::percent) +
  labs(title = "Mood Disorders", y = "Number of Participants", subtitle = "Participants per self-declared mood disorder", x="Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

p2 <- select(df, Participant, Gender, starts_with("DisorderTreatment_")) |> 
  pivot_longer(cols = starts_with("DisorderTreatment_"), names_to = "Treatment", values_to = "Value") |> 
  mutate(Treatment = str_remove_all(Treatment, fixed("DisorderTreatment_"))) |> 
  summarize(N = sum(Value) / nrow(df), .by=c("Gender", "Treatment")) |> 
  mutate(N_tot = sum(N), .by="Treatment") |>
  mutate(Treatment = fct_reorder(Treatment, desc(N_tot))) |>
  ggplot(aes(x = Treatment, y = N, fill=Gender)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Male"= "#64B5F6", "Female"= "#F06292", "Other"="orange", "Missing"="brown")) +
  scale_y_continuous(expand = c(0, 0), labels=scales::percent) +
  labs(title = "Treatment", y = "Number of Participants", subtitle = "Participants per treatment", x="Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1))

@@

df$Depression <- rowSums(select(df, DisorderTreatment_MoodStabilizer, DisorderTreatment_Antidepressant))
df$Depression <- ifelse(df$Depression > 0, TRUE, FALSE)

@@

`r sum(df$MoodDisorder)` participants (`r insight::format_percent(sum(df$MoodDisorder) / nrow(df))`) have a mood disorder, as indexed by the use of a mood stabilizer, antidepressant, or anxiolytic treatment. `r sum(df$Depression)` participants (`r insight::format_percent(sum(df$Depression) / nrow(df))`) have a mood disorder, as indexed by the use of a mood stabilizer or antidepressant. `r sum(df$DisorderHistory, na.rm=TRUE)` participants (`r insight::format_percent(sum(df$DisorderHistory, na.rm=TRUE) / nrow(df))`) reported having a history of mood disorder.

>>>>

---
title: "PHQ-4-R - Data Cleaning (Study 2)"
editor: source
editor_options: 
  chunk_output_type: console
format:
  html:
    code-fold: true
    self-contained: false
---

@@

We removed `r length(outliers$attentionchecks)` (`r insight::format_percent(length(outliers$attentionchecks) / nrow(df))`) participants for failing the attention checks.

@@

### Response Coherence and Variability

@@

df$BDI2_SD <- select(df, matches("BDI2_"), -BDI2_Duration, -BDI2_Order, -BDI2_Total) |> 
  SD_per_dim(dims=c("BDI2_")) |> 
  rowMeans()
df$STAI5_SD <- select(df, matches("STAI5_"), -STAI5_Duration, -STAI5_Order, -STAI5_General) |> 
  SD_per_dim(dims=c("STAI5_")) |> 
  rowMeans()
df$IAS_SD <- select(df, matches("IAS_"), -IAS_Duration, -IAS_Order, -IAS_Total) |> 
  SD_per_dim(dims=c("IAS_")) |> 
  rowMeans()
```

```{r}
m <- mgcv::gam(Failed ~ s(IAS_SD), 
          data=mutate(df, Failed = ifelse(AttentionCheck == "Failed", 1, 0)), 
          family="binomial") 
parameters::parameters(m)

estimate_relation(m, length=40) |> 
  ggplot(aes(x=IAS_SD, y=Predicted)) +
  geom_ribbon(aes(ymin=CI_low, ymax=CI_high), alpha=0.2) +
  geom_line() + 
  ggside::geom_xsidedensity(data=df, aes(fill=AttentionCheck), alpha=0.3, color=NA)

df |> 
  select(Participant, IAS_SD, AttentionCheck) |> 
  mutate(Participant = fct_reorder(Participant, IAS_SD)) |>
  ggplot(aes(x=Participant, y=IAS_SD, fill=AttentionCheck)) +
  geom_bar(stat="identity") +
  theme_minimal() +
  theme(axis.text.x = element_blank())

outliers$ias_sd <- as.character(df[abs(standardize(df$IAS_SD, robust=TRUE)) > qnorm(0.995), "Participant"])
outliers$ias_sd <- outliers$ias_sd[!outliers$ias_sd %in% c(outliers$attentionchecks, outliers$duration)]
```

We removed `r length(outliers$ias_sd)` (`r insight::format_percent(length(outliers$ias_sd) / nrow(df))`) participants for failing being an outlier on the IAS' SD value.

@@

```{r}
dfoutlier <- df |> 
  select(contains("IAS_"), -IAS_Duration, -IAS_Order, -IAS_Total, -IAS_SD) |>
  # select(contains("BDI2_"), -BDI2_Duration, -BDI2_Order, -BDI2_Total, -BDI2_SD) |>
  # select(contains("STAI5_"), -STAI5_Duration, -STAI5_Order, -STAI5_General, -STAI5_SD) |>
  # select(contains("MAIA2_"), -MAIA_Duration, -MAIA_Order) |>
  # datawizard::remove_empty_rows() |> 
  performance::check_outliers(
    method=c("optics", "mahalanobis_robust"),
    threshold=list(optics=2, optics_xi=0.1)) |> 
  as.data.frame() |> 
  mutate(Participant = fct_reorder(df$Participant, Distance_OPTICS),
         Outlier_AttentionCheck = ifelse(Participant %in% outliers$attentionchecks, 1, 0),
         Outlier_Distance = ifelse(abs(standardize(Distance_OPTICS, robust=TRUE)) > qnorm(0.995), 1, 0),
         Experiment_Duration = df$Experiment_Duration,
         Outlier = ifelse(Outlier_AttentionCheck == 1, "Failed Attention Checks", "Passed"))

outliers$distance <- as.character(dfoutlier[dfoutlier$Outlier_Distance == 1, "Participant"])
outliers$distance <- outliers$distance[!outliers$distance %in% c(outliers$attentionchecks, outliers$duration, outliers$ias_sd)]

@@

We removed `r length(outliers$distance)` (`r insight::format_percent(length(outliers$distance) / nrow(df))`) participants based on their multivariate distance. `r length(c(outliers$ias_sd, outliers$distance))` (`r insight::format_percent(length(c(outliers$ias_sd, outliers$distance)) / nrow(df))`) outliers in total.

```{r}
df <- filter(df, !Participant %in% c(outliers$attentionchecks, outliers$ias_sd, outliers$distance))
df <- select(df, -starts_with("AttentionCheck"))
```

@@

```{r}
p1 <- select(df, Participant, Gender, starts_with("Disorder_")) |> 
  pivot_longer(cols = starts_with("Disorder_"), names_to = "Disorder", values_to = "Value") |> 
  mutate(Disorder = str_remove_all(Disorder, fixed("Disorder_"))) |> 
  summarize(N = sum(Value) / nrow(df), .by=c("Gender", "Disorder")) |> 
  mutate(N_tot = sum(N), .by="Disorder") |>
  mutate(Disorder = fct_reorder(Disorder, desc(N_tot))) |>
  ggplot(aes(x = Disorder, y = N, fill=Gender)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Male"= "#64B5F6", "Female"= "#F06292", "Other"="orange", "Missing"="brown")) +
  scale_y_continuous(expand = c(0, 0), labels=scales::percent) +
  labs(title = "Mood Disorders", y = "Number of Participants", subtitle = "Participants per self-declared mood disorder", x="Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal()

p2 <- select(df, Participant, Gender, starts_with("DisorderTreatment_")) |> 
  pivot_longer(cols = starts_with("DisorderTreatment_"), names_to = "Treatment", values_to = "Value") |> 
  mutate(Treatment = str_remove_all(Treatment, fixed("DisorderTreatment_"))) |> 
  summarize(N = sum(Value) / nrow(df), .by=c("Gender", "Treatment")) |> 
  mutate(N_tot = sum(N), .by="Treatment") |>
  mutate(Treatment = fct_reorder(Treatment, desc(N_tot))) |>
  ggplot(aes(x = Treatment, y = N, fill=Gender)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = c("Male"= "#64B5F6", "Female"= "#F06292", "Other"="orange", "Missing"="brown")) +
  scale_y_continuous(expand = c(0, 0), labels=scales::percent) +
  labs(title = "Treatment", y = "Number of Participants", subtitle = "Participants per treatment", x="Type") +
  theme(axis.text.x = element_text(angle = 45, hjust = 1)) +
  theme_minimal()

@@

df$Treatment <- rowSums(select(df, DisorderTreatment_MoodStabilizer, DisorderTreatment_Antidepressant, DisorderTreatment_Anxiolytic, DisorderTreatment_Therapy))
df$Treatment <- ifelse(df$Treatment > 0, TRUE, FALSE)

df$Depression <- rowSums(select(df, Disorder_MDD, Disorder_Dysthymia))
df$Depression <- ifelse(df$Depression > 0 & df$Treatment, TRUE, FALSE)

df$Anxiety <- rowSums(select(df, Disorder_GAD, Disorder_Panic))
df$Anxiety <- ifelse(df$Anxiety > 0 & df$Treatment, TRUE, FALSE)

@@

- `r sum(df$MoodDisorder)` participants (`r insight::format_percent(sum(df$MoodDisorder) / nrow(df))`) have a mood disorder, as indexed by the use of a mood stabilizer, antidepressant, or anxiolytic treatment. 
- `r sum(df$Depression)` participants (`r insight::format_percent(sum(df$Depression) / nrow(df))`) were labelled as having Depression, as indexed by the self-reported presence of MDD together with the use of a treatment (antidepressent, anxiolytic and/or therapy). 
- `r sum(df$Anxiety)` participants (`r insight::format_percent(sum(df$Anxiety) / nrow(df))`) were labelled as having Anxiety, as indexed by the self-reported presence of GAD or Panic Disorder together with the use of a treatment (antidepressent, anxiolytic and/or therapy).
- `r sum(df$DisorderHistory, na.rm=TRUE)` participants (`r insight::format_percent(sum(df$DisorderHistory, na.rm=TRUE) / nrow(df))`) reported having a history of mood disorder.
